Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-18 aistats.tex
Enhance multivariate MLE proof and add it to aistats

>>>>

\subsection{Multivariate MLE}
In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, the mean parameter $\mu_*$ becomes a positive definite covariance matrix of order $d$, with $p = \frac{d(d+1)}{2} $ degrees of freedom.
Its entropy is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, with gradient negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$, and associated Bregman divergence
\alignn{
	\bregmanconj(\mu_*; \hat \mu_n) 
	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
}
Thanks to the linearity of the trace, the expectation becomes
\alignn{
	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
}

When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
then the mean parameter "ratio" is
\aligns{
	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
}
but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$. 
We know formulas for the expectation of an inverse Wishart and the log-determinant of a Wishart (thanks to $\log\det$ being a sufficient statistics and $\nabla\logpart$)
\aligns{
	\E[\mV] &= n \mI \\
	\forall n > d+1,\E[\mV^{-1}] &= \frac{\mI}{n - d - 1} 
	\implies \Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}  \\
	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
Putting it all together, we get the following closed form for the expectation of the KL
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}

The complicated term is the $\psi_d$ function.
We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
	\label{eq:psid_bound}
\end{align}
We can bound the harmonic sum thanks to the typical bound
\aligns{
	\inv{2n +1}
	\leq H_n - \log(n) - \gamma \leq 
	\inv{2n - 1}
}
where $\gamma$ is the Euler constant. Then
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
\end{align}
Plugging this back into~\eqref{eq:psid_bound} yields
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
\end{align}
where we emphasized that the sum runs up to $d$.
Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n}
\end{align}
which yields the final rate
\begin{align}
 	\boxed{
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} \;.
 	}
\end{align}
where $p = d(d+1)/2$.
This rate is asymptotically tight.

### 2021-10-18 main.tex
Enhance multivariate MLE proof and add it to aistats

\subsection{Multivariate MLE}
In higher dimensions $d$
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
Using the same upper bound as in the univariate case
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{d-1} \log(1 - \frac{i}{n}) - \frac{1}{n-i} \; .
\end{align}
We can bound the harmonic sum with typical bounds
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
\end{align}
which yields 
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}  \; .
\end{align}
From there we have two options. 
Either we use a concave tangent bound on the logarithms to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n}
\end{align}
which yields the final rate
\begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{d(d+1)}{4n} + \frac{d(d+1)^2 + \half[d+1]}{2 n (n - d -1)} \;.
\end{align}
Note that this rate is not tight for $d>1$, but it is tight for $d=1$, and we almost recover the previous rate.

Another option, potentially tighter than the tangent bound, is to bound the sum of logarithms with an integral, using the monotonicity of the logarithm $\log(1 - \frac{i}{n}) \leq \log(1- \frac{x}{n}) , \forall x \in [i - 1, i]$
\begin{align}
	\sum_{i=1}^{d} \log(1 - \frac{i}{n})
	&\leq \int_{0}^{d} \log(1- \frac{x}{n}) dx \\
	& = n \int_{1 - \frac{d}{n}}^{1} \log(y) dy \\
	& = n [ y (\log(y) - 1) ]_{1 - \frac{d}{n}}^{1} \\
	& =  - n - (n -d) (\log ( 1 - \frac{d}{n}) - 1) \\
	& = - (n -d) \log ( 1 - \frac{d}{n}) - d \; .
\end{align}
Unfortunately, it is hard to get a tractable and non-vacuous bound for this guy. At $n=d$ it is worth $-d$, and then it grows up to $0$. and we can bound it using some exponents of $1 - \frac{d}{n}$, but that is not really helpful to get an easy formula.
Actually looking at \href{https://www.desmos.com/calculator/q4sqk73tld}{graphics}, the first upper bound turns out to be tighter than the second one.

>>>>

\subsection{Multivariate MLE}
In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, the mean parameter $\mu_*$ becomes a positive definite covariance matrix of order $d$, with $p = \frac{d(d+1)}{2} $ degrees of freedom.
Its entropy is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, with gradient negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$, and associated Bregman divergence
\alignn{
	\bregmanconj(\mu_*; \hat \mu_n) 
	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
}
Thanks to the linearity of the trace, the expectation becomes
\alignn{
	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
}

When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
then the mean parameter "ratio" is
\aligns{
	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
}
but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$. 
We know formulas for the expectation of an inverse Wishart and the log-determinant of a Wishart (thanks to $\log\det$ being a sufficient statistics and $\nabla\logpart$)
\aligns{
	\E[\mV] &= n \mI \\
	\forall n > d+1,\E[\mV^{-1}] &= \frac{\mI}{n - d - 1} 
	\implies \Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}  \\
	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
Putting it all together, we get the following closed form for the expectation of the KL
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}

The complicated term is the $\psi_d$ function.
We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
	\label{eq:psid_bound}
\end{align}
We can bound the harmonic sum thanks to the typical bound
\aligns{
	\inv{2n +1}
	\leq H_n - \log(n) - \gamma \leq 
	\inv{2n - 1}
}
where $\gamma$ is the Euler constant. Then
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
\end{align}
Plugging this back into~\eqref{eq:psid_bound} yields
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
\end{align}
where we emphasized that the sum runs up to $d$.
Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n}
\end{align}
which yields the final rate
\begin{align}
 	\boxed{
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} \;.
 	}
\end{align}
where $p = d(d+1)/2$.
This rate is asymptotically tight.
